Changes in accessibility to foodbanks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario

### 2021-04-23 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
Add to accessibility analysis

% Pandoc header
\usepackage[table]{xcolor}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

In Canada, the number of foodbanks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of foodbanks
and their use comes from the non-profit association Foodbanks Canada,
which conducts an annual assessment through its affiliated members.
Foodbanks Canada's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member foodbanks across the
country, and found that Canadians visited foodbanks 1.1 million times in
March of 2018. Of those accessing foodbanks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to Foodbanks
Canada's 2018 data, single-adult households represent 45\% of those
utilizing foodbanks despite making up 28\% of Canada's population, 19\%
are single-parent households (compared to 10\% nationally), and 35\% of
those accessing foodbank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing foodbanks list social or
disability assistance as their primary source of income. Moreover,
children aged 0-18 are over-represented as foodbank users, making up
35\% of those accessing foodbank services despite representing 20\% of
the country's population. Similarly, using data from the 2011-2012 CCHS,
Tarasuk et al. (2019) found higher odds of food insecurity amongst
households relying on social assistance, those without a university
degree or with children under the age of 18, and individuals that lived
alone, renters, and those identifying as Aboriginal. While surveys
revealed that only 20 to 30 percent of those experiencing food
insecurity were found to frequent foodbanks in Canada (Tarasuk et al.,
2014), research from Ottawa (Enns et al., 2020) and Vancouver (Black and
Seto, 2020) suggests that long-term users tend to be older, have health
or mobility challenges, live in large households, and are less likely to
have employment income.

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from Foodbanks Canada showed that 52\% of member foodbanks reported an
increase in usage in March of 2020 when initial lockdown restrictions
were put in place across much of the country. The pandemic also created
significant staffing issues with 42\% of foodbanks reporting a reduction
in volunteers. However, 53\% of foodbanks reported a decrease in use
into the summer of 2020 which foodbanks attributed to emergency
financial support programs from the federal government were.
Nevertheless, some of these benefit programs were temporary, which
suggests that many households may again turn to foodbank services to
meet their needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing foodbanks, the locational or
transportation accessibility aspect of foodbank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socio-economic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For foodbanks specifically, previous qualitative
research in Ontario has noted difficulties associated with
transportation for trips to the foodbank (Smith-Carrier et al., 2017).
For example, Smith et al. (2017) note that participants access foodbanks
using a variety of modes of transportation with respondent stating ``I
wish it {[}the foodbank{]} was a little more centrally located. Because
if I didn't have a bike I'd have to walk it all the way out there and
back. I wonder about people who don't.'' From this, the authors conclude
that ``transportation can be challenging, particularly if the foodbank
is situated in a remote location.'' To offer greater insight into the
role of transportation and location in foodbank accessibility, this
research examines how geographic accessibility to foodbanks and food
services changed in Hamilton during the COVID-19 pandemic.

>>>>

% Pandoc header
\usepackage[table]{xcolor}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

@@

In Canada, the number of foodbanks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of foodbanks
and their use comes from the non-profit association Food Banks Canada
(FBC), which conducts an annual assessment through its affiliated
members. FBC's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member foodbanks across the
country, and found that Canadians visited foodbanks 1.1 million times in
March of 2018. Of those accessing foodbanks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to FBC's 2018
data, single-adult households represent 45\% of those utilizing
foodbanks despite making up 28\% of Canada's population, 19\% are
single-parent households (compared to 10\% nationally), and 35\% of
those accessing foodbank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing foodbanks list social or
disability assistance as their primary source of income. Similarly,
using data from the 2011-2012 CCHS, Tarasuk et al. (2019) found higher
odds of food insecurity amongst households relying on social assistance,
those without a university degree or with children under the age of 18,
and individuals that lived alone, renters, and those identifying as
Aboriginal. While surveys revealed that only 20 to 30 percent of those
experiencing food insecurity were found to frequent foodbanks in Canada
(Tarasuk et al., 2014), research from Ottawa (Enns et al., 2020) and
Vancouver (Black and Seto, 2020) suggests that long-term users tend to
be older, have health or mobility challenges, live in large households,
and are less likely to have employment income.

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from FBC showed that 52\% of member foodbanks reported an increase in
usage in March of 2020 when initial lockdown restrictions were put in
place across much of the country. The pandemic also created significant
staffing issues with 42\% of foodbanks reporting a reduction in
volunteers. However, 53\% of foodbanks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary, which suggests that many
households may again turn to foodbank services to meet their needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing foodbanks, the locational or
transportation accessibility aspect of foodbank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socio-economic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For foodbanks specifically, previous qualitative
research by Smith et al. (2017) in Ontario has noted that participants
access foodbanks using a variety of modes of transportation and that
some food bank users have expressed difficulties with transportation for
trips to foodbank locations, with one respondent stating ``I wish it
{[}the foodbank{]} was a little more centrally located. Because if I
didn't have a bike I'd have to walk it all the way out there and back. I
wonder about people who don't.'' From this, the authors conclude that
``transportation can be challenging, particularly if the foodbank is
situated in a remote location.'' To offer greater insight into the role
of transportation and location in foodbank accessibility, this research
examines how geographic accessibility to foodbanks and food services
changed in Hamilton during the COVID-19 pandemic.
